---
title: 'CLL Cytokine Screen 2021: Figure 1'
author: "Holly Giles and Peter Bruch"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
      toc: yes
      toc_depth: 3
      toc_float: yes
      code_folding: "hide" 
---

# Figure 1

In this sub-vignette we present the analysis and source code for figure 1. This sub-vignette can be built along with all other sub-vignettes by running CLLCytokineScreen2021.Rmd. 


## Set up 
Load libraries
```{r loadLibraries1, cache = FALSE, message = FALSE, warning = FALSE }

library(gridExtra)
library(ggplot2)
library(magick)
library(patchwork)
library(cowplot)
library(dplyr)
library(tidyr)
library(tidyverse)

```

Set plot directory
```{r plotDir1}
plotDir = ifelse(exists(".standalone"), "", "../../inst/figs/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
```


## Load data  
```{r loadData1}

#drugs: dataframe containing meta data on all drugs used in screen 
load( "../../data/drugs.RData")

#cytokines: dataframe containing meta data on all stimuli used in screen 
load( "../../data/cytokines.RData")

#df: tibblecontaining all screening viability data
load( "../../data/df.RData")

```


```{r loadData_fromtsv1, eval = FALSE}

#from tsvs
df <- read.table(file= "../../inst/extdata/df.txt",header = TRUE) %>% as_tibble()
drugs <- read.delim("../../inst/extdata/Drugs.txt")
cytokines <- read.delim("../../inst/extdata/Cytokines.txt")

df$Cytokine <- factor(df$Cytokine, levels= c("No Cytokine",
                                              "Resiquimod", 
                                              "IL-4", 
                                              "TGF-b1",
                                              "IL-1b",
                                              "Interferon gamma",
                                              "SDF-1a",
                                              "sCD40L" ,
                                              "sCD40L+IL-4",
                                              "soluble anti-IgM", 
                                              "CpG ODN",
                                              "IL-6",
                                              "IL-10",
                                              "IL-21",
                                              "HS-5 CM", 
                                              "IL-15",
                                              "BAFF",
                                              "IL-2" ))

df$Drug <- factor(df$Drug, levels =c("DMSO",
                                     "BAY-11-7085",
                                     "Everolimus",
                                     "Fludarabine",
                                     "I-BET 762",
                                     "Ibrutinib","Idelalisib",
                                     "Luminespib",
                                     "Nutlin-3a",
                                     "PRT062607",
                                     "Pyridone-6",
                                     "Ralimetinib",
                                     "Selumetinib"))


```


## Define Aesthetics
```{r defineAesthetics1}

source("../../R/themes_colors.R")

```

## Plot figures
### Fig 1A     
Plot graphic of screen overview
```{r Fig1A, fig.width=5, fig.height=5}

#plot pdf file
Fig1A  <- magick::image_read_pdf("../../inst/images/Fig1A.pdf", pages = 1, density = 100)

Fig1A <- ggdraw() + draw_image(Fig1A, 
                               scale = 1)

Fig1A
```


### Fig 1B
Overview of cytokine pathways
```{r Fig1B, fig.width=8, fig.height = 5}

#plot pdf file
Fig1B  <- magick::image_read_pdf("../../inst/images/Fig1B.pdf", pages = 1, density = 100)

Fig1B <- ggdraw() + draw_image(Fig1B, 
                               scale = 1)
Fig1B

```

### Fig 1C
Overview of drug pathways and approval status
```{r Fig1C, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=4, fig.height=3.5}

#set table theme
ttable <- ttheme_minimal(
  core=list(fg_params=list(col = darkergrey,fontface=3)),
  
  colhead=list(fg_params=list(col="black", fontface=4L),
               bg_params = list(fill= colors[1])))
#make table
Fig1C<-
  drugs %>% 
  dplyr::select(`Name`, `main_targets`, `target_category`) %>%
  dplyr::rename(Drug=`Name`, Target=`main_targets`, Category=`target_category`) %>% 
  dplyr::arrange(Category) %>% 
  gridExtra::tableGrob(theme = ttable, rows=NULL) 


wrap_elements(Fig1C)

```


### Fig 1D
```{r Fig1D,   fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.height=7, fig.width=10, warning = FALSE}
# Data preparation
#:::::::::::::::::::::::::::::::::::::::::::

mydata<-df%>%
  filter((Drug=="DMSO"&Cytokine!="No Cytokine")|(Drug!="DMSO"&Cytokine=="No Cytokine"), Drug_Concentration%in%c("High", "None"))%>%
  group_by(PatientID, Drug, Cytokine) %>%
  summarize(Log=mean(Log), .groups = "keep") %>%
  ungroup() %>%
  mutate(Cytokine=as.character(Cytokine), Drug=as.character(Drug)) %>%
  mutate(Condition=case_when(Cytokine=="No Cytokine"~ Drug, Drug=="DMSO"~Cytokine)) %>%
  select(PatientID, Condition, Log)%>%
  spread(Condition, Log)%>%
  ungroup() %>% 
  column_to_rownames("PatientID")

# Compute correlation matrix
#::::::::::::::::::::::::::::::::::::::::::
# Correlation matrix between all variables
cor.mat <- mydata %>% rstatix::cor_mat()


Cor.p.value<-cor.mat %>% rstatix::cor_get_pval() %>% 
  column_to_rownames( var="rowname")%>%
  as_tibble(rownames="Component_1") %>%
    dplyr::mutate(Component_1=factor(Component_1)) %>%
  pivot_longer(-Component_1, names_to="Component_2", values_to="p_value")

cor.mat <- column_to_rownames(cor.mat, var="rowname")  

# Define Lists of Stimuli and Drugs
Cytokines <-unique(df$Cytokine) %>% setdiff("No Cytokine")
Drugs<-unique(df$Drug) %>% setdiff("DMSO")

# Define Theme for plots
t_tmp<-theme(
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.major.x = element_blank(),
  panel.grid.minor = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.line = element_line(),
  axis.text.x  = element_text(size = 12, face="bold"),
  axis.text.y = element_text(size = 12, face="bold"),
  axis.ticks.x = element_blank())


#Drug
Druglist<-c("Pyridone-6",  "BAY-11-7085", "Ralimetinib", "Everolimus",  "Selumetinib", "Idelalisib",  "Ibrutinib",   "PRT062607",   "Fludarabine", "Nutlin-3a",   "I-BET 762",   "Luminespib")

Drug_data_tmp<-cor.mat[Druglist, Druglist]

Drug_Corr<-Drug_data_tmp

    Drug_Corr[upper.tri(Drug_Corr, diag = TRUE)]<- NA
    
    
# Drug_heatmap_clusters<-pheatmap::pheatmap(Drug_data_tmp)    


# Drug_data_order<-rownames(Drug_data_tmp[Drug_heatmap_clusters$tree_row[["order"]],])

Drug_Corr_tab<-Drug_Corr %>%
  as_tibble(rownames="Drug_1") %>%
  pivot_longer(-Drug_1, names_to="Drug_2", values_to="Pearson_R") %>%
  dplyr::filter(!is.na(Pearson_R)) %>% 
  left_join(Cor.p.value, by=c("Drug_1"="Component_1", "Drug_2"="Component_2")) %>% 
  mutate(adj.p.value=p.adjust(p_value, method="BH"))
    
Drug_Cor_Plot<- Drug_Corr_tab %>% 
    dplyr::mutate(Drug_1=factor(Drug_1, levels = rev(Druglist)),
                   Drug_2=factor(Drug_2, levels = (Druglist))) %>%
  ggplot(aes(Drug_1, Drug_2))+
  geom_tile(aes(fill=Pearson_R),color = "grey")+
  scale_fill_gradientn(colors=c("#003DA5",  "white",  "#A6093D"), limits=c(-1,1))+
  geom_text(aes(label=ifelse(adj.p.value<0.05,sub("^(-?)0.", "\\1.", sprintf("%.2f", round(Pearson_R,2))),""), fontface = ifelse(adj.p.value<0.05, 1, 1)), size=4)+
    t_tmp+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))+
    labs(fill = "Pearson R", x="", y="")+
  guides(fill="none")

Fig1D <- Drug_Cor_Plot
Fig1D

```

### Fig 1E
```{r Fig1E,   fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.height=7, fig.width=10, warning=FALSE}

# Cytokine

tmp_heatmap<-pheatmap::pheatmap(cor.mat[which(rownames(cor.mat)%in%Cytokines), which(colnames(cor.mat)%in%Cytokines)], silent = TRUE)

Cytokinelist<-rownames(cor.mat[which(rownames(cor.mat)%in%Cytokines), which(colnames(cor.mat)%in%Cytokines)][tmp_heatmap$tree_row[["order"]],])

Cyt_Corr<-cor.mat[Cytokinelist, Cytokinelist]


    Cyt_Corr[upper.tri(Cyt_Corr, diag = TRUE)]<- NA

Cyt_Corr_tab<-Cyt_Corr %>%
  as_tibble(rownames="Cyt_1") %>%
    # dplyr::mutate(Cyt_1=factor(Cyt_1, levels = rev(rownames(Cyt_Corr)))) %>%
  pivot_longer(-Cyt_1, names_to="Cyt_2", values_to="Pearson_R") %>%
  dplyr::filter(!is.na(Pearson_R)) %>% 
  left_join(Cor.p.value, by=c("Cyt_1"="Component_1", "Cyt_2"="Component_2")) %>% 
  mutate(adj.p.value=p.adjust(p_value, method="BH"))    
    
Cyt_Cor_Plot<-  Cyt_Corr_tab %>%
    dplyr::mutate(Cyt_1=factor(Cyt_1, levels = rev(Cytokinelist)),
                  Cyt_2=factor(Cyt_2, levels = (Cytokinelist))) %>%
  ggplot(aes(Cyt_1, Cyt_2))+
  geom_tile(aes(fill=Pearson_R),color = "grey")+
  scale_fill_gradientn(colors=c("#003DA5",  "white",  "#A6093D"), limits=c(-1,1))+
  geom_text(aes(label=ifelse(adj.p.value<0.05,sub("^(-?)0.", "\\1.", sprintf("%.2f", round(Pearson_R,2))),""), fontface = ifelse(adj.p.value<0.05, 1,1)), size=4)+
    t_tmp+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))+
    labs(fill = "Pearson R", x="", y="")+
  scale_y_discrete(labels=c("TGF-b1"="TGF-\u03B21", "sCD40L+IL-4"="sCD40L + IL4", "IL-1b"="IL1\u03B2", "IL-4"="IL4", "IL-6"="IL6","IL-15"="IL15","IL-10"="IL10", "IL-21"="IL21","IL-2"="IL2", "Interferon gamma"= "Interferon \u03B3", "SDF-1a"="SDF-1\u03B1"))+
  scale_x_discrete(labels=c("TGF-b1"="TGF-\u03B21", "sCD40L+IL-4"="sCD40L + IL4", "IL-1b"="IL1\u03B2", "IL-4"="IL4", "IL-6"="IL6","IL-15"="IL15","IL-10"="IL10", "IL-21"="IL21","IL-2"="IL2", "Interferon gamma"= "Interferon \u03B3", "SDF-1a"="SDF-1\u03B1"))+
  guides(fill="none")

Fig1E <- Cyt_Cor_Plot
Fig1E

```




## Assemble Figure
```{r, Figure1, fig.height=11, fig.width=14, fig.path=plotDir, dev=c("png", "cairo_pdf"), message=FALSE, warning=FALSE}

tp <- theme(plot.tag=element_text(size = 30, face="plain"))


design1 <-"
  ABBC
  DDEE
"

Figure1 <-
  
  wrap_elements(Fig1A) + tp+
  wrap_elements(Fig1B) + tp+
  wrap_elements(Fig1C ) + tp+
  wrap_elements(Fig1D) + tp+
  wrap_elements(Fig1E) + tp+
  
  plot_annotation(tag_levels = "A", title="Figure 1", theme = theme(title=element_text(size = 20))) +
  plot_layout(design = design1, heights = c(0.7,1.2), width=c(1,.5,1,1.1))


Figure1


```


## Appendix
```{r appendix1}
Sys.info()
sessionInfo()
```
