---
title: 'CLL Cytokine Screen 2021: Figure 7'
author: "Holly Giles and Peter Bruch"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
      toc: yes
      toc_depth: 3
      toc_float: yes
      code_folding: "hide" 
---

# Figure 7 
In this sub-vignette we present the analysis and source code for figure 7. This sub-vignette can be built along with all other sub-vignettes by running CLLCytokineScreen2021.Rmd. 

## Set up
Load libraries 
```{r loadLibraries7, cache = FALSE, message = FALSE, warning = FALSE}

library(patchwork)
library(maxstat)
library(survminer)
library(survival)
library(ggpubr)
library(rlang)
library(ggbeeswarm)
library(dplyr)

```
 
Set plot directory
```{r plotDir7}
plotDir = ifelse(exists(".standalone"), "", "../../inst/figs/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
```


## Load data
Raw
```{r loadData7}

#ihc_patient_data: 596 x 7 tibble containing IHC staining intensities for pIRAK4, pSTAT6 and STAT6 staining in CLL and healthy tissue 
load( "../../data/ihc_patient_data.RData")

#ihc_surv_data: 200 x 15 tibble containing IHC staining intensities for pIRAK4, pSTAT6 and STAT6 staining in CLL and healthy tissue, along with meta data (including clinical outcomes) on each patient sample
load( "../../data/ihc_surv_data.RData")

```


```{r loadData_fromtsv7, eval = FALSE}

#ihc_patient_data: 596 x 7 tibble containing IHC staining intensities forpIRAK4, pSTAT6 and STAT6 staining in CLL and healthy tissue 
ihc_patient_data <- read.table(file= "../../inst/extdata/ihc_patient_data.txt",header = TRUE) %>% as_tibble() %>% dplyr::group_by(PatientID)

#ihc_surv_data: 200 x 15 tibble containing IHC staining intensities for pIRAK4, pSTAT6 and STAT6 staining in CLL and healthy tissue, along with meta data (including clinical outcomes) on each patient sample
ihc_surv_data <- read.table(file= "../../inst/extdata/ihc_surv_data.txt",header = TRUE) %>% as_tibble() %>% dplyr::group_by(PatientID)


```


Process
```{r processData7}

#convert stain type to factor
ihc_patient_data$Stain <- factor(ihc_patient_data$Stain, levels = c("pSTAT6", "STAT6" , "pIRAK4"))

ihc_patient_data<-filter(ihc_patient_data, !is.na(Diagnosis))

```


## Set aesthetics
```{r defineAesthetics7}

source("../../R/themes_colors.R")

```


## Plot Figures
### Figure 7A
Beeswarm-boxplot of mean staining intensities of pSTAT6, for healthy and CLL LN samples
```{r, Fig7A, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=5, fig.height=7}
plotTab<-
  ihc_patient_data %>%
  #get staining intensities for pSTAT6 only
  dplyr::filter(Stain %in% c("pSTAT6"))
 
Fig7A <- 
  #plot staining intensity, stratified by CLL / Healthy LN
  ggplot(plotTab, aes(x = Tissue, y = Intensity)) +
  geom_boxplot() +
  geom_beeswarm(aes(color=Tissue), alpha=1, cex=2) +
  #add p values
  stat_compare_means(method = "t.test", comparisons=list(c(1,2)), size=6) +
  scale_color_manual(values = c(palreds[8], palblues[2])) +
  xlab("") +
  ylab("Mean pSTAT6 Staining Intensity") +
  guides(color = "none") +
  scale_x_discrete(labels = c( "CLL"="CLL-infiltrated \nlymph nodes", "LK"="Non-neoplastic \nlymph nodes")) +
  coord_cartesian(clip = "off") +
  #add preset theme 2
  t2

Fig7A


plotTab %>% 
  group_by(Tissue) %>% 
  count()
  
```

### Figure 7B
Beeswarm-boxplot of mean staining intensities of pIRAK4, for healthy and CLL LN samples
```{r Fig7B, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=7, fig.height=7}

Fig7B <-
  ihc_patient_data %>%
  #filter for pIRAK4 staining intensities only
  dplyr::filter(Stain%in%c("pIRAK4")) %>%
  #plot staining intensity, stratified by CLL / Healthy LN
  ggplot(aes(x=Tissue, y=Intensity)) +
  geom_boxplot() +
  geom_beeswarm(aes(color=Tissue), alpha=1, cex=2) +
  #add p values 
  stat_compare_means(method = "t.test", comparisons=list(c(1,2)), size=6) +
  scale_color_manual(values = c(palreds[8], palblues[2])) +
  xlab("") +
  ylab("Mean pIRAK4 Staining Intensity") +
  guides(color="none") +
  scale_x_discrete(labels = c( "CLL"="CLL-infiltrated \nlymph nodes", "LK"="Non-neoplastic \nlymph nodes")) +
  coord_cartesian(clip = "off") +
  #add preset theme 2
  t2



Fig7B

 ihc_patient_data %>%
  #filter for pIRAK4 staining intensities only
  dplyr::filter(Stain%in%c("pIRAK4"))  %>% 
  group_by(Diagnosis) %>% 
  count()

```

### Figure 7C
Image of IHC cross-section of CLL-infiltrated LN stained for pSTAT6
```{r Fig7C, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=6, fig.height=4}

Fig7C <-cowplot::ggdraw() +
  cowplot::draw_image( "../../inst/images/CLL1_I3_pSTAT6_scale.png", scale = 1)

Fig7C

```



### Figure 7D
Image of IHC cross-section of healthy LN stained for pSTAT6
```{r Fig7D, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=6, fig.height=4}

##read image
Fig7D <- cowplot::ggdraw() + 
  cowplot::draw_image("../../inst/images/LK2_B4_pSTAT6_scale.png", scale = 1)

Fig7D

```

### Figure 7E
Image of IHC cross-section of CLL-infiltrated LN stained for pIRAK4
```{r Fig7E, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=6, fig.height=4}

Fig7E <- cowplot::ggdraw() + 
  cowplot::draw_image( "../../inst/images/CLL1_I3_pIRAK4_scale.png", scale = 1)

Fig7E

```


### Figure 7F
Image of IHC cross-section of healthy LN stained for pIRAK4
```{r Fig7F, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=6, fig.height=4}


Fig7F <- cowplot::ggdraw() + 
  cowplot::draw_image( "../../inst/images/LK2_B4_pIRAK4_scale.png", scale = 1)

Fig7F
```


### Figure 7G
Kaplan-Meier curve to show associations of pSTAT6 levels with treatment free survival.     
In Figures 7G and 7H the two pSTAT6 and pIRAK4 groups (high /low) were defined by  mean staining intensities dichotomised using maximally selected rank statistics. The same 64 CLL lymph node samples were used for both Kaplan-Meier plots. 50 patient samples were in the high pSTAT6 group, and 52 in the high pIRAK4 group.       

Prework for 7G and H
```{r KaplanMeier_prework7, warning=FALSE}

#Define stains for which want to visualize TTT
stains <- c("pSTAT6", "pIRAK4")


#Get optimal cut offs 
stats <- lapply(stains, function(stn){
 
  survival <- dplyr::select(ihc_surv_data, PatientID, Tissue, Diagnosis, Sex, TTT, treatedAfter, stn)
  colnames(survival) <- c("PatientID", "Tissue", "Diagnosis", "Sex", "TTT", "treatedAfter", "target")
  
  #Run test to obtain cut off threshold for high pSTAT6 versus low pSTAT6

  maxtest <- maxstat.test(Surv(TTT, treatedAfter)~ target, 
                          data = survival,
                          smethod = "LogRank",
                          alpha = NULL)
  
  cutpoint <- maxtest$estimate
  
 })
  
  names(stats) <- c("pSTAT6", "pIRAK4")
  
                                          
  #Annotate by cutoff point
  survdf <- mutate(ihc_surv_data, 
                   phosphoSTAT6 = ifelse(pSTAT6 < stats$pSTAT6, "low", "high"), 
                   phosphoIRAK4 = ifelse(pIRAK4 < stats$pIRAK4, "low", "high"))
  
                                      

  #fit survival models
  f_pstat6 <- survfit(Surv(TTT, treatedAfter) ~ phosphoSTAT6, data= survdf)
  f_pirak4 <- survfit(Surv(TTT, treatedAfter) ~ phosphoIRAK4, data= survdf)
  
  fits <- list(pSTAT6 = f_pstat6, pIRAK4 = f_pirak4)
  
  #Make plot
  gg = ggsurvplot_list(fits, 
                       survdf,  
                       pval=TRUE, 
                       palette=c(palreds[8],palreds[3]),  
                       risk.table = TRUE, legend.title = "", 
                       ggtheme = t2, 
                       legend.labs =list(c("High", "Low"),c("High", "Low")),  
                       xlab="Time in years", ylab="\nTime to next treatment (probability)", title= "", legend = "bottom")

  

```


```{r, Fig7G, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=16, fig.height=7}

#get plot for pSTAT6 stain
 Fig7G <-
  wrap_elements(gg$pSTAT6$plot + 
                  theme(axis.title.x = element_blank()) +
                  gg$pSTAT6$table + 
                  theme(plot.title = element_blank()) +
                  plot_layout(ncol = 1, heights = c(85, 15)))
  
Fig7G

```


### Figure 7H
Kaplan-Meier curve to show associations of pSTAT6 levels with treatment free survival.   
```{r Fig7H, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=16, fig.height=7}

#get plot for pIRAK4 stain 
Fig7H <- wrap_elements(gg$pIRAK4$plot+ 
                         theme(axis.title.x = element_blank()) +
                         gg$pIRAK4$table + 
                         theme(plot.title = element_blank()) +
                  plot_layout(ncol=1, heights = c(85, 15))) 


Fig7H

```


## Assemble Figure
```{r, Figure7, fig.path=plotDir, dev=c("png", "pdf", "jpeg"),fig.height=15, fig.width=16, dpi=300}

design1 <-"
  ACG
  ADG
  BEH
  BFH
  "
tp <- theme(plot.tag = element_text(size = 30, vjust = 1, face="plain"))

Fig7 <-
  wrap_elements(Fig7A) + tp +
  wrap_elements(Fig7B) + tp +
  Fig7C + tp +
  Fig7D + tp +
  Fig7E + tp +
  Fig7F + tp +
  
  Fig7G + tp +
  
  Fig7H + tp +
  
  plot_layout(design = design1, widths = c(1,0.7,1), heights =c(0.8, 1, 0.8, 1))+
  plot_annotation(tag_levels = "A", title="Figure 7", theme = theme(title=element_text(size = 20)))
  
Fig7

```


## Count tables
```{r countTables}

ihc_surv_data %>% 
  ungroup() %>% 
  mutate(TFT = ifelse(is.na(TFT)|TFT == "NA", "NA", "Available" )) %>% 
  dplyr::group_by(Diagnosis, TFT) %>% 
  count()

```

## Appendix
```{r appendix7}
Sys.info()

sessionInfo()
```

