---
title: 'CLL Cytokine Screen 2021: Figure 5'
author: "Holly Giles & Peter Bruch"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: "hide" 
---

# Figure 5 

In this sub-vignette we present the analysis and source code for figure 5.  This sub-vignette can be built along with all other sub-vignettes, by running CLLCytokineScreen2021.Rmd. 

## Set up  
Load Libraries 
```{r loadLibraries5, cache = FALSE, message = FALSE, warning = FALSE}

library(patchwork)
library(ggplot2)
library(data.table)
library(magrittr)
library(ggplot2)
library(pheatmap)
library(tidyr)
library(dplyr)
```

Set plot directory
```{r plotDir5}
plotDir = ifelse(exists(".standalone"), "", "../../inst/figs/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
```


## Load data
Raw
```{r loadData5}

#df: tibble containing all screening viability data
load( "../../data/df.RData")

```


```{r loadData_fromtsv5, eval = FALSE}

#from tsvs
df <- read.table(file= "../../inst/extdata/df.txt",header = TRUE) %>% as_tibble()

df$Cytokine <- factor(df$Cytokine, levels= c("No Cytokine",
                                              "Resiquimod", 
                                              "IL-4", 
                                              "TGF-b1",
                                              "IL-1b",
                                              "Interferon gamma",
                                              "SDF-1a",
                                              "sCD40L" ,
                                              "sCD40L+IL-4",
                                              "soluble anti-IgM", 
                                              "CpG ODN",
                                              "IL-6",
                                              "IL-10",
                                              "IL-21",
                                              "HS-5 CM", 
                                              "IL-15",
                                              "BAFF",
                                              "IL-2" ))

df$Drug <- factor(df$Drug, levels =c("DMSO",
                                     "BAY-11-7085",
                                     "Everolimus",
                                     "Fludarabine",
                                     "I-BET 762",
                                     "Ibrutinib","Idelalisib",
                                     "Luminespib",
                                     "Nutlin-3a",
                                     "PRT062607",
                                     "Pyridone-6",
                                     "Ralimetinib",
                                     "Selumetinib"))


```



Process data
```{r processData5}
# List of Cytokines, Drugs and treatments 
  ##Drugs
  thedrugs <- unique(df$Drug) 

  ##Cytokines
  thecytokines <- unique(df$Cytokine) 

  #Treatments
  #drugs
  drugtreatments <- list()
  drugtreatments <- 
    lapply(thedrugs, function(x){
      paste("treatment_drug", x, sep='')
      })
  
  
  #cytokines
  cytokinetreatments <- list()
  cytokinetreatments <- 
    lapply(thecytokines, function(y){
      paste("treatment_cytokine", y, sep='')
      })


# Tidy up data frame to use in linear model
df <- 
  dplyr::filter(df,
                #use high concentrations only
                Drug_Concentration %in% c("High","None")) %>% 
  #select columns for linear model
  dplyr::select(PatientID, Drug, Cytokine, Log) %>%  
        
  #rename columns 
  plyr::rename(c("Drug"="treatment_drug", "Cytokine"="treatment_cytokine", "Log"="Viability"))

```


## Set aesthetics
```{r defineAesthetics5}

source("../../R/themes_colors.R")

```

## Pre - work
Fit the linear model
```{r linearModel5, message=FALSE}


# define the base level per treatment_type as the "no"-treatment
df$treatment_drug <- as.factor(df$treatment_drug)
df$treatment_cytokine <- as.factor(df$treatment_cytokine)

df$treatment_drug %<>% relevel("DMSO")
df$treatment_cytokine %<>% relevel("No Cytokine")


# inspect design matrix of interaction model
# model.matrix(~treatment_drug + treatment_cytokine + treatment_drug:treatment_cytokine, df)

# fit model with interaction
fit <- lm(Viability ~ treatment_drug * treatment_cytokine, df)


```

Extract p values
```{r getPvalues5, warning = FALSE}

#extract p values

pvalues <- summary(fit)$coefficients[,4]

#order as a dataframe
pvaldf <- as.data.frame(as.matrix(pvalues)) %>% 
          setDT(keep.rownames = TRUE)
          
#rename columns
colnames(pvaldf) <- c("treatment", "pvalue")

#filter out single agent treatments
singletreatments <- c(drugtreatments, cytokinetreatments, "(Intercept)")

pvaldf <- dplyr::filter(pvaldf, !treatment %in% singletreatments)


pvaldf <- 
  pvaldf %>% 
  #remove "treatment" prefix
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_drug", "", .)))) %>% 
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_cytokine", "", .))))
  
  #split drug:cytokine into two columns
pvaldf <-   
  data.frame(pvaldf, do.call(rbind, strsplit(pvaldf$treatment, split = ":", fixed = TRUE)))

#renove treatment column
pvaldf <- pvaldf[,c("pvalue","X1", "X2")]

#rename columns
colnames(pvaldf) <- c("pvalue","drug", "Cytokine")



```


## Plot Figures
### Fig 5A
```{r, Fig5A, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=8, fig.height=8, warning = FALSE}

dummydata <-
  as.data.frame(list(c(1,2,3,4),
                     c(0,0,0,0),
                     c(-0.6,-0.6,0,-0.1), 
                     c(0.6,0.6,0.1,-0.1), 
                     c(0.5,-0.5,0.8,-0.8)), 
                col.names=c("plot",
                            "coord1", 
                            "coord2", 
                            "coord3", 
                            "coord4")) %>% 
  as_tibble()%>% 
  
  mutate(expected = coord2 + coord3, measured = coord4) %>% 
  
  pivot_longer(cols = coord1:measured, names_to = c("plot"), values_to="y.value", names_repair="unique" ) %>% 
  
  dplyr::rename(Plot = plot...1, x.value = plot...2) %>% 
  
  mutate(x.value = factor(x.value, levels = c("coord1", "coord2", "coord3", "coord4", "expected", "measured"))) %>% 
  
  #add annotation of interaction type
  mutate(Nature=if_else(Plot%in%c(1,2),"Antagonistic", "Synergistic"),
         Nature=factor(Nature, levels = c("Antagonistic", "Synergistic")),
         Direction=if_else(Plot%in%c(1,3), "Positive \u03B2 int", "Negative \u03B2 int"),
         Direction=factor(Direction, levels = c("Positive \u03B2 int", "Negative \u03B2 int")),
         Annotation=case_when(Plot==1~"I",
                              Plot==2~"II",
                              Plot==3~"III",
                              Plot==4~"IV"),
         Plot=as.character(Plot)

         )

Fig5A <-
      ggplot(dplyr::filter(dummydata, !x.value%in%c("expected", "measured")), 
             aes(x=x.value, y=y.value, group=Plot, color=Plot))+
  
  geom_hline(yintercept = 0, linetype=2, color="black")+
  geom_point(size=1)+
  geom_line(size=1)+
  geom_text(aes(label=Annotation), x=1, y=.8,  size=7)+ 
  scale_color_manual(values = c("#A6093D","#003DA5","#A6093D","#003DA5"))+
  guides(color="none")+      

  geom_segment(data=dplyr::filter(dummydata, grepl('expected', x.value)), aes(x = 3.5, y = y.value, xend = 4.5, yend = y.value), color="blue", size=2) +
  geom_segment(data=dplyr::filter(dummydata, grepl('measured', x.value)), aes(x = 3.5, y = y.value, xend = 4.5, yend = y.value), color="black", size=2)+
  
  theme_bw() +
  t1 +
  xlab("") +
  ylab("Logarithm of relative viability") +
  coord_cartesian(ylim = c(-1, 1)) +
  scale_y_continuous(n.breaks=3)+
  scale_x_discrete(labels=c( "coord1"="DMSO","coord2"="Drug","coord3"="Stimulus","coord4"="Drug +\nStimulus")) +
    theme(axis.ticks.x= element_blank(),axis.ticks.y = element_blank(),
          plot.tag=element_text(size = 20)) +
        facet_grid(Nature~Direction)


Fig5A

```



### Fig 5B
Barplot quantifying interaction types 
```{r, Fig5B,  fig.path=plotDir, dev=c("png", "cairo_pdf"),fig.width = 3, fig.height = 6}


fit_as_df <- 
  fit$coefficients %>% 
  as.matrix() %>% 
  as.data.frame() %>%
  setDT(keep.rownames = TRUE) %>% 
  as_tibble() %>% 
  dplyr::rename(Condition = rn, comb_coefficient = V1) %>% 
  dplyr::filter(grepl("drug", Condition)&grepl("cytokine", Condition)) %>% 
  mutate(Condition = gsub("treatment_drug", "", Condition)) %>% 
  mutate(Condition = gsub("treatment_cytokine", "", Condition)) %>% 
  separate(Condition, c("Drug", "Cytokine"), sep=":")

drug_fit_as_df <- 
  fit$coefficients %>% 
  as.matrix() %>%
  as.data.frame() %>%
  setDT(keep.rownames = TRUE) %>% 
  as_tibble() %>% 
  dplyr::rename(Drug = rn, drug_coefficent=V1) %>% 
  dplyr::filter(grepl("drug", Drug)&!grepl("cytokine", Drug)) %>% 
  mutate(Drug = gsub("treatment_drug", "", Drug))


cyt_fit_as_df <-
  fit$coefficients %>% 
  as.matrix() %>% 
  as.data.frame() %>%
  setDT(keep.rownames = TRUE) %>% 
  as_tibble() %>% 
  dplyr::rename(Cytokine=rn, cyt_coefficent=V1) %>% 
  dplyr::filter(!grepl("drug", Cytokine)&grepl("cytokine", Cytokine)) %>% 
  mutate(Cytokine = gsub("treatment_cytokine", "", Cytokine))


fit_combinations <-
  left_join(fit_as_df, drug_fit_as_df, by = "Drug") %>% 
  left_join(cyt_fit_as_df, by = "Cytokine") %>% 
  mutate(predicted_coefficient=cyt_coefficent+drug_coefficent) %>% 
  mutate(obs_coef=predicted_coefficient+comb_coefficient) %>%
  left_join(mutate(pvaldf,drug=as.character(drug),Cytokine=as.character(Cytokine)), by=c("Drug"="drug", "Cytokine"="Cytokine")) %>% 
  dplyr::filter(pvalue<=0.05) %>%
  mutate(Category_Symbol=case_when(
    comb_coefficient>=0&(obs_coef<cyt_coefficent|obs_coef<drug_coefficent)~ "I",  
    comb_coefficient>=0&(obs_coef>cyt_coefficent&obs_coef>drug_coefficent)~ "III",
    comb_coefficient<0&(obs_coef>cyt_coefficent|obs_coef>drug_coefficent)~ "II",  
    comb_coefficient<0&(obs_coef<cyt_coefficent&obs_coef<drug_coefficent)~ "IV"
    )) %>% 
  mutate(Category=case_when(
    Category_Symbol=="I"~ "Positive Antagonistic",  
    Category_Symbol=="III"~ "Positive Synergistic",
    Category_Symbol=="II"~ "Negative Antagonistic",  
    Category_Symbol=="IV"~ "Negative Synergistic"
    )) %>% 
  mutate( Category=factor(Category, levels=c("Positive Antagonistic","Negative Antagonistic","Positive Synergistic","Negative Synergistic")))
  

Fig5B <-
  fit_combinations %>% 
  dplyr::group_by(Category) %>% 
  tally() %>% 
  
ggplot(aes(x=Category, y=n, fill=Category))+
  geom_bar(width = 1, stat = "identity")+
  t1+
  guides(fill="none")+
  xlab("")+ ylab("")+
  scale_fill_manual(values = c("#A6093D","#003DA5","#A6093D","#003DA5"))

Fig5B
```


### Fig 5C
Heatmap of significant drug : stimulus interaction coefficients    
Get interaction coefficients matrix for plotting heatmap 
```{R getBeta5}


#extract beta interaction values
betavalues <- summary(fit)$coefficients[,1]

#create matrix of beta values 
betadf <- as.data.frame(as.matrix(betavalues)) %>%
          setDT(keep.rownames = TRUE)

#rename columns
colnames(betadf) <- c("treatment", "betavalue")

#remove beta values for drugs and cytokines alone
betadf <- dplyr::filter(betadf, !treatment %in% singletreatments)

#remove treatment prefix
betadf <- betadf %>% 
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_drug", "", .)))) %>% 
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_cytokine", "", .))))

#split drug:cytokine by colon
betadf <- data.frame(betadf, do.call(rbind, strsplit(betadf$treatment, split = ":", fixed = TRUE)))

#select columns of interest
betadf <- betadf[,c("betavalue","X1", "X2")]

#rename columns 
colnames(betadf) <- c("betavalue","drug", "Cytokine")


```


```{r Fig5C,  fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width=7.5, fig.height = 7, warning=FALSE}

#set significance
a <- 0.05

#make a matrix of beta values, where  beta = 0 if corresponding p val is > than significance threshold 
pvalmat <- xtabs(pvalue~drug+Cytokine, data=pvaldf)
bvalmat <- xtabs(betavalue~drug+Cytokine, data=betadf)

#check in same order
# bvalmat[,colnames(pvalmat)] 
# bvalmat[rownames(pvalmat),]

bvalmat[pvalmat >= a] = 0

#transform matrix 
bvalmat <- t(bvalmat)

tree <- 
  pheatmap(bvalmat, silent = TRUE)


cyt_order <- rownames(bvalmat[tree$tree_row[["order"]],])

drug_order <- colnames(bvalmat[,tree$tree_col[["order"]]])


Fig5C <-
  #append p values
  left_join(betadf, pvaldf, by = c("drug", "Cytokine")) %>% 
  #make beta value 0 if p value is < sig
  mutate(betavalue = ifelse(pvalue<a, betavalue, 0)) %>% 
  #put drugs in order of clustered heatmap
  mutate(drug=factor(drug, levels = drug_order)) %>% 
  
  #put stimuli in order of clustered heatmap
  mutate(Cytokine=factor(Cytokine, levels = rev(cyt_order))) %>% 
  
  #make heatmap with ggplot
  ggplot(aes(x=drug, y=Cytokine))+
  
  geom_tile(aes(fill=betavalue),color = "grey")+
  #set colour scale
  scale_fill_gradientn(colors=c("#003DA5",  "white",  "#A6093D"), limits=c(-0.6,0.6))+
  #set theme
  t2+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.x =element_blank(),
        panel.background = element_blank(), 
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        legend.key.height=unit(2.5, "cm"))+
  labs(fill = expression(paste(beta["int"], "-value")))+
  #add category
  geom_text(data = fit_combinations, aes(x=Drug, label=Category_Symbol) )+
  scale_y_discrete(labels=c("TGF-b1"="TGF-\u03B21", "sCD40L+IL-4"="sCD40L + IL4", "IL-1b"="IL1\u03B2", "IL-4"="IL4", "IL-6"="IL6","IL-15"="IL15","IL-10"="IL10", "IL-21"="IL21","IL-2"="IL2", "Interferon gamma"= "Interferon \u03B3", "SDF-1a"="SDF-1\u03B1"))
  
Fig5C

```

### Fig 5D - G

Preparation: Line plots of examples of drug and stimulus combinations that show an interaction
```{r FigD-1}

#get lists of drugs and cytokines without baseline treatments
thedrugs.only <- thedrugs %>% setdiff("DMSO") %>% sort()
thecytokines.only <- thecytokines %>% setdiff("No Cytokine")%>% sort()

# get a list of interactions for which p value of beta interaction is significant
sign_conditions <-
  pvaldf %>% 
  dplyr::filter(pvalue<=0.05) %>% 
  mutate(condition = paste0("treatment_drug:", 
                            drug, 
                            ":treatment_cytokine:",
                            Cytokine))


gg <- vector(mode="list", length=length(sign_conditions$condition))
names(gg) <- sign_conditions$condition



plotList <- 
lapply(names(gg), function(i){
  
  #preparations for the interaction plot
  drug <- dplyr::filter(sign_conditions, condition ==i)$drug
  drugTreat <- paste("treatment_drug", drug, sep='')
      
  cyt <- dplyr::filter(sign_conditions, condition ==i)$Cytokine
  cytokineTreat <-paste("treatment_cytokine", cyt, sep='')
      
      
  Category <- fit_combinations[which(fit_combinations$Drug==drug &
                               fit_combinations$Cytokine==cyt),]$Category
  
  #extract coefficients to plot

  ## baseline effect
  baseline <- fit$coefficients["(Intercept)"]

  ## single treatment effect for type_1 = "a"
  single_1 <- fit$coefficients[drugTreat]

    ## single treatment effect for type_2 = "b"
    single_2 <- fit$coefficients[cytokineTreat]

    # interaction Viability
    interaction <- fit$coefficients[paste(drugTreat, cytokineTreat, sep=':')]

      
    ## observed single treatment values
    observed_single_1 <- baseline + single_1 
    observed_single_2 <- baseline + single_2

    ## observed double treatment value
    observed_double <- baseline + single_1 + single_2 + interaction

    ## predicted double treatment value (based only on single effects)
    predicted_double <- baseline + single_1 + single_2

    xlabels <- c("DMSO", drug, 
                 case_when(cyt=="IL-4"~"IL4",
                           cyt=="Interferon gamma"~"Interferon \u03B3",
                           cyt=="sCD40L+IL-4"~"sCD40L + IL4",
                           cyt=="soluble anti-IgM"~"soluble\nanti-IgM",
                           TRUE~cyt),
                paste(drug, " +\n",
                       case_when(cyt=="IL-4"~"IL4",
                           cyt=="Interferon gamma"~"Interferon \u03B3",
                           cyt=="sCD40L+IL-4"~"sCD40L + IL4",
                           TRUE~cyt),
                sep=""))
      
    segment_size <- 2
    
    
    plotTab <-
      df %>%
      
      dplyr::filter(treatment_drug %in% c("DMSO" , drug)) %>%

      dplyr::filter(treatment_cytokine %in% c("No Cytokine", cyt)) %>%

      dplyr::mutate(treatment_combination = 
                  #if baseline treatment
                  ifelse(treatment_drug=="DMSO" & 
                         treatment_cytokine=="No Cytokine",
                         "DMSO",
                         #if drug only treatment
                         ifelse(treatment_drug == drug & 
                                treatment_cytokine =="No Cytokine",
                                paste0("Drug=", drug),
                                #if cytokine only treatment
                                ifelse(treatment_drug == "DMSO" & 
                                       treatment_cytokine == cyt,
                                       paste0("Cytokine=", cyt),
                                       #otherwise combinatorial treatment
                                       paste0("Drug=", drug, "\nCytokine=", cyt))))) %>%
  #make treatment_combination as a factor
  mutate(treatment_combination = factor(treatment_combination, 
                                        levels=c("DMSO", 
                                                 paste0("Drug=", drug),
                                                 paste0("Cytokine=", cyt),
                                                 paste0("Drug=", drug, "\nCytokine=", cyt)
                                                 )))
  #make plot
        
        

  ggplot(plotTab, aes(treatment_combination, Viability, group=PatientID)) +
  
  geom_hline(yintercept = 0, linetype=2, color="black")+

  lemon::geom_pointline( size=1,   
                         colour=ifelse(interaction > 0,
                                       "#A6093D","#003DA5"), alpha=0.3) +
  #add predicted viability with control treatment
  geom_segment( size=segment_size,
                aes(x=1-.2, xend=1+.2, y=baseline, yend=baseline),
                colour = "black") +
  #add predicted viability with drug only treatment
  geom_segment( size=segment_size,
                aes(x=2-.2, xend=2+.2, y=observed_single_1,yend=observed_single_1),
                colour="black") +
  #add predicted viability with cytokine only treatment
  geom_segment( size=segment_size,
                aes(x=3-.2, xend=3+.2, y=observed_single_2, yend=observed_single_2),
                colour="black") +
  #add predicted viability with both treatments
  geom_segment( size=segment_size,
                aes(x=4-.2, xend=4+.2, y=observed_double, yend=observed_double),
                colour="black") +
  #add predicted viability with both treatments, without interaction
  geom_segment( size=segment_size,
                aes(x=4-.2, xend=4+.2, y=predicted_double, 
                    yend=predicted_double), color="blue") + 
 
   #add category and combination as title
    
    
    
  ggtitle(paste( Category,": ", drug, " & ", 
                 case_when(cyt=="IL-4"~"IL4",
                           cyt=="Interferon gamma"~"Interferon \u03B3",
                           cyt=="sCD40L+IL-4"~"sCD40L + IL4",
                           TRUE~cyt), sep="")) + 
  
  #add theme
  t2 + theme(axis.text.x = element_text(size=fontsize+5)) +
    theme(plot.tag=element_text(size = 30))+
  scale_x_discrete(labels = xlabels) +
  ylab("Logarithm of relative viability") + 
  xlab("")


  } ) 

names(plotList) <- names(gg)


```


#### Fig 5D
```{r Fig5D,  fig.path=plotDir, dev=c("png", "cairo_pdf"), warning=FALSE}

wrap_plots(plotList$`treatment_drug:Ibrutinib:treatment_cytokine:IL-4`)

```

#### Fig 5E
```{r Fig5E,  fig.path=plotDir, dev=c("png", "cairo_pdf"), warning=FALSE}

wrap_plots(plotList$`treatment_drug:Ibrutinib:treatment_cytokine:Interferon gamma`)
  
```

#### Fig 5F
```{r Fig5F,  fig.path=plotDir, dev=c("png", "cairo_pdf"), warning=FALSE}
  
wrap_plots(plotList$`treatment_drug:Pyridone-6:treatment_cytokine:sCD40L+IL-4`)

```

#### Fig 5G
```{r Fig5G,  fig.path=plotDir, dev=c("png", "cairo_pdf"), warning=FALSE}

wrap_plots(plotList$`treatment_drug:Ralimetinib:treatment_cytokine:Interferon gamma`)

```

#### Fig 5H
```{r Fig5H,  fig.path=plotDir, dev=c("png", "cairo_pdf"), warning=FALSE}

wrap_plots(plotList$`treatment_drug:Ibrutinib:treatment_cytokine:CpG ODN`)
```

#### Fig 5G
```{r Fig5I,  fig.path=plotDir, dev=c("png", "cairo_pdf"), warning=FALSE}

wrap_plots(plotList$`treatment_drug:Luminespib:treatment_cytokine:soluble anti-IgM`)

```

## Arrange plots
```{r, Figure5, fig.path=plotDir, dev=c("png", "cairo_pdf"),  fig.width=20, fig.height=22, warning=FALSE}

design1 <-"
  1223
  4455
  6677
  8899
  "
Fig5<-
wrap_elements(Fig5A) + theme(plot.tag=element_text(size = 30)) +

wrap_elements(Fig5B) + theme(plot.tag=element_text(size = 30)) +
  
wrap_elements(Fig5C) + theme(plot.tag=element_text(size = 30)) +
  
wrap_plots(plotList$`treatment_drug:Ibrutinib:treatment_cytokine:IL-4`)+
  
wrap_plots(plotList$`treatment_drug:Ibrutinib:treatment_cytokine:Interferon gamma`)+
  
wrap_plots(plotList$`treatment_drug:Pyridone-6:treatment_cytokine:sCD40L+IL-4`)+
  
wrap_plots(plotList$`treatment_drug:Ralimetinib:treatment_cytokine:Interferon gamma`)+
  
wrap_plots(plotList$`treatment_drug:Ibrutinib:treatment_cytokine:CpG ODN`)+
  
wrap_plots(plotList$`treatment_drug:Luminespib:treatment_cytokine:soluble anti-IgM`)+
  
plot_layout(design=design1,
            heights = c(1,0.5,0.5,0.5),
            widths = c(0.92,0.08,0.05,0.95)) +

  plot_annotation( tag_levels = "A", title="Figure 5", theme = theme(title=element_text(size = 20)))

Fig5

```

## Appendix
```{r appendix5}
Sys.info()
sessionInfo()
```

                