---
title: 'CLL Cytokine Screen 2021: Figure 4'
author: "Holly Giles and Peter Bruch"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
      toc: yes
      toc_depth: 3
      toc_float: yes
      code_folding: "hide" 
---

# Figure 4
In this sub-vignette we present the analysis and source code for figure 4.  This sub-vignette can be built along with all other sub-vignettes, by running CLLCytokineScreen2021.Rmd. 


## Set up 
```{r setup4}

set.seed(1996)

```

Load libraries
```{r loadLibraries4, cache = FALSE, message = FALSE, warning = FALSE}

library(plyr)
library(clusterProfiler)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(ChIPseeker)
library(genomation)
library(ggbeeswarm)
library(ggpubr)
library(ggplot2)
library(patchwork)
library(msigdbr)
library(png)
library(cowplot)
library(tidyverse)
library(Hmisc)
library(pheatmap)
library(RColorBrewer)
library(org.Hs.eg.db)


```

Set plot directory
```{r plotDir4, message=FALSE}
plotDir = ifelse(exists(".standalone"), "", "../../inst/figs/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
```


## Load data
Load raw files
```{r loadData4, message=FALSE}

#df: tibble containing all screening viability data
load( "../../data/df.RData")

#patMeta: tibble containing all patient genetic data
load( "../../data/patMeta.RData")


#diffTF_large: tibble containing diffTF weighted mean difference values  for trisomy 12 versus non-trisomy 12 CLL ATACseq data
load( "../../data/diffTF_large.RData")

#diffTF_ibet: tibble containing diffTF weighted mean difference values  for 4 ibet versus 4 control treated CLL ATACseq samples
load( "../../data/diffTF_ibet.RData")


## diffTF conversion tables: 
#HOCOMOCO IDs for trisomy 12 versus non-trisomy 12 diffTF run 
tri12_TFs <- read.table(file= "../../inst/extdata/translationTable_hg19.csv",header = TRUE) %>% as_tibble()

#HOCOMOCO IDs for IBET-762 versus control diffTF run 
ibet_TFs <- read.table(file= "../../inst/extdata/translationTable_hg38.csv",header = TRUE) %>% as_tibble()

```

```{r loadData_fromtsv, eval= FALSE}

#df: tibble containing all screening viability data
df <- read.table(file= "../../inst/extdata/df.txt",header = TRUE) %>% as_tibble()

#patMeta: tibble containing all patient genetic data
patMeta <- read.table(file= "../../inst/extdata/patMeta.txt",header = TRUE) %>% as_tibble()


#diffTF_large: tibble containing diffTF weighted mean difference values for 636 TFs, for trisomy 12 versus non-trisomy 12 CLL ATACseq data
diffTF_large <- read_tsv("../../inst/extdata/BIGCLL52.permutations.summary.tsv.gz")

#diffTF_IBET: tibble containing diffTF weighted mean difference values for 636 TFs, for trisomy 12 versus non-trisomy 12 CLL ATACseq data
diffTF_ibet <- read_tsv("../../inst/extdata/IBETvsCTRL.summary.tsv.gz")


#Arrangedata
#df
df$Cytokine <- factor(df$Cytokine, levels= c("No Cytokine",
                                              "Resiquimod", 
                                              "IL-4", 
                                              "TGF-b1",
                                              "IL-1b",
                                              "Interferon gamma",
                                              "SDF-1a",
                                              "sCD40L" ,
                                              "sCD40L+IL-4",
                                              "soluble anti-IgM", 
                                              "CpG ODN",
                                              "IL-6",
                                              "IL-10",
                                              "IL-21",
                                              "HS-5 CM", 
                                              "IL-15",
                                              "BAFF",
                                              "IL-2" ))

#patMeta

patMeta <- patMeta %>%mutate_at(vars(-PatientID), as.factor)

```

Process files
```{r processData4}  

df_complete <- df

#get Cytokine only data
df <- dplyr::filter(df, 
                    Drug == "DMSO", 
                    Cytokine != "No Cytokine")
   
```


## Define Aesthetics
```{r defineAesthetics4}

source("../../R/themes_colors.R")

```

## Plot Figures
### Figure 4A
TLR Response stratified by IGHV and trisomy 12
```{r Fig4A, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.height = 5.5, fig.width = 6}

#filter for Resiquimod-only treatment, only show patients who are annotated for IGHV and Trisomy12
plotTab <- left_join(df, patMeta, by = "PatientID") %>% 
  dplyr::filter(Cytokine=="Resiquimod",
                !is.na(trisomy12),
                !is.na(IGHV.status))

Fig4A =  
 
  plotTab %>%
         
  ggplot(aes(x=interaction(trisomy12, IGHV.status),
                    y=Log,
                    color=(IGHV.status)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  scale_x_discrete(labels=c("0.M"="IGHV-M\n WT",
                            "0.U"="IGHV-U\n WT",
                            "1.M"="IGHV-M\n trisomy 12",
                            "1.U"="IGHV-U\n trisomy 12"))+
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     comparisons = list(c(3,4),c(1,3),c(1,2)),
                     size=5)+
  xlab("") +
  ylab("Log(Viability)") +
  ggtitle("Resiquimod (TLR)") +
  scale_color_manual(values=c(palblues[3], palreds[8])) + 
  coord_cartesian(ylim = c(-1.5, 3.1), clip="off") +
  t2
  # theme(axis.text.x = element_text(angle = 45, vjust =1))


rm(plotTab)

Fig4A


```

### Figure 4B
```{r Fig4B, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.height = 5.5, fig.width = 6}

#filter for IL4-only treatment, only show patients who are annotated for Trisomy12
plotTab <- left_join(df, patMeta, by = "PatientID") %>% 
  dplyr::filter(Cytokine=="IL-4",
                !is.na(trisomy12))

Fig4B =  
 
  plotTab %>%
         
  ggplot(aes(x=interaction(trisomy12),
                    y=Log,
                    color=(trisomy12)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  scale_x_discrete(labels=c("0"="WT",
                            "1"="trisomy 12"))+
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     comparisons = list(c(1,2)),
                     size=5)+
  xlab("") +
  ylab("Log(Viability)") +
  coord_cartesian( clip="off") +
  ggtitle("IL4") +
  scale_color_manual(values=c(palblues[3], palreds[8])) + 
  t2


rm(plotTab)

Fig4B


```


### Figure 4C
TGFbeta response stratified by trisomy 12
```{r, Fig4C, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.height = 5.5, fig.width = 6, warning = FALSE}

#filter for TGFbeta-only treatment, only show patients who are annotated for IGHV and Trisomy12
plotTab <- left_join(df, patMeta, by = "PatientID") %>% 
  dplyr::filter(Cytokine=="TGF-b1",
                !is.na(trisomy12))

Fig4C =  
 
  plotTab %>%
         
  ggplot(aes(x=trisomy12,
                    y=Log,
                    color=(trisomy12)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
 
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     comparisons = list(c(1,2)),
                     size=5)+
  scale_x_discrete(labels=c("0"="WT",
                            "1"="trisomy 12"))+
  xlab("") +
  ylab("Log(Viability)") +
  ggtitle("TGF\u03B2") +
  scale_color_manual(values=c(palblues[3], palreds[8])) + 

  t2

rm(plotTab)

Fig4C


```


### Figure 4D
Trisomy 12 versus WT ATACseq
Comparison of TF binding site accessibility shows that Spi-B and other TFs show higher accessibility of their binding sites in trisomy 12 in 2 independent datasets.
```{r, Fig4D, fig.path = plotDir, dev = c("png", "cairo_pdf"), fig.height = 6, fig.width=15}

#Get all TFs that show differential activity
plotTab.diffTF <- dplyr::filter(diffTF_large, pvalueAdj < 0.05)

#change to name of TF quoted in text
plotTab.diffTF$TF <- gsub("SPI1", "PU1", plotTab.diffTF$TF)

#order by weighted mean difference
idx <- order(plotTab.diffTF[["weighted_meanDifference"]], decreasing = TRUE)
plotTab.diffTF$TF <- factor(plotTab.diffTF$TF,levels=rev(unique(plotTab.diffTF$TF[idx])))


#plot figure
Fig4D <-
  ggplot(plotTab.diffTF,
         aes(x=weighted_meanDifference,
             y=TF,
             fill= ifelse(weighted_meanDifference>0,
                          palreds[5], palblues[1]))) +
  geom_bar(stat = "identity", width = 0.75) +
  xlim(c(-0.2, 0.175)) +

  #colours and theme
  scale_fill_manual(values = c(palblues[1], palreds[8]), guide = "none") +
  t1 +
  theme(axis.text.x = element_text(angle = 45, vjust =1))+

  #Labels
  ylab("") +
  xlab("Change in inferred TF activity") +
  ggtitle("Diffential TF binding site accessibility in trisomy 12 CLL") +

  #Add annotations
  #arrow 1
  annotate("segment", x = -0.15, xend = -0.15, y = 8.4, yend = 0.75,
           colour = "#5e5e5e", size=3, alpha=1, arrow=arrow()) +
  #arrow 2
  annotate("segment", x = -0.15, xend = -0.15, y = 8.6, yend = 17,
           colour = "#5e5e5e", size=3, alpha=1, arrow=arrow()) +

   #2 arrow labels
  annotate("text", x = c(-0.185,-0.185), y = c(4.8,12.5),
          label = c("Lower in trisomy 12 CLL",
                    "Higher in trisomy 12 CLL") ,
          color="black",
          size=5 , fontface="bold") +

  #flip plot
  coord_flip()


Fig4D

```

### Figure 4E
IBET-762 response stratified by trisomy 12
```{r, Fig4E, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.height = 5, fig.width = 5}

Fig4E <- 
  left_join(df_complete, patMeta, by = "PatientID") %>% 
  dplyr::filter(Drug=="I-BET 762", Drug_Concentration == "High", Cytokine == "No Cytokine") %>%
      filter(!is.na(trisomy12)) %>%
         
  ggplot(aes(x = trisomy12, y = Log, color = (trisomy12)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     size=5,
                     comparisons = list(c(1,2)))+
  ylab("Log(Viability)") +
  ggtitle("IBET-762") +
  scale_color_manual(values=c(colors[1], colors[2])) + 
  coord_cartesian(ylim = c(-1.25, 0.6), clip="off") +

  t2+
  scale_x_discrete(labels=c("0"="WT",
                            "1"="trisomy 12"))+
  xlab("")

Fig4E

```


### Figure 4F
Comparison of TF binding site accessibility in trisomy 12 vs non-trisomy 12 samples and in IBET vs control treated samples

```{r, Fig4F, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.width = 12, fig.height=10}

colnames(ibet_TFs) <- c("Symbol", "ENSEMBL", "TF")

#add ensembl IDs to diffTF_ibet
diffTF.ibet <- left_join(diffTF.ibet, ibet_TFs[2:3], by = "TF")
colnames(diffTF.ibet)[1] <- "TF_original"

#add TF names from tri12 run to diffTF_ibet
colnames(tri12_TFs)[3] <- "TF"
diffTF.ibet <- left_join(diffTF.ibet, tri12_TFs[2:3], by = "ENSEMBL")

#change to name of TF quoted in text
diffTF.ibet$TF <- gsub("SPI1", "PU1", diffTF.ibet$TF)

#filter for tri 12 signature TFs
plotTab.diffTF.ibet <- diffTF.ibet %>% filter(TF %in%  plotTab.diffTF$TF)

#order by weighted mean difference
idx <- order(plotTab.diffTF[["weighted_meanDifference"]], decreasing = TRUE)
plotTab.diffTF.ibet$TF <- factor(plotTab.diffTF.ibet$TF,levels=rev(unique(plotTab.diffTF$TF[idx])))


Fig4F <-
  ggplot(plotTab.diffTF.ibet,
         aes(x=weighted_meanDifference,
             y=TF,
             fill= ifelse(weighted_meanDifference>0,
                          palreds[5], palblues[1]))) +
  geom_bar(stat = "identity", width = 0.75) +
  xlim(c(-0.2, 0.175)) +

  #colours and theme
  scale_fill_manual(values = c(palblues[1], palreds[8]), guide = "none") +
  t1 +
  theme(axis.text.x = element_text(angle = 45, vjust =1))+

  #Labels
  ylab("") +
  xlab("Change in inferred TF activity") +
  ggtitle("Diffential TF binding site accessibility in IBET-762-treated CLL") +



  #flip plot
  coord_flip()


Fig4F


```



## Assemble Figure 4


```{r Fig4, fig.path=plotDir, dev=c("png", "cairo_pdf"), fig.height=18, fig.width=14, eval = TRUE, warning = FALSE}

design1<-"
  ABC
  DDE
  FFG
"

tp <- theme(plot.tag=element_text(size = 30, face="plain"))

Fig4 <-


 wrap_elements( Fig4A )+ tp +
 wrap_elements( Fig4B )+ tp +
 wrap_elements( Fig4C )+ tp + 
 wrap_elements( Fig4D )+ tp + 
 wrap_elements( Fig4E )+ tp +
 wrap_elements( Fig4F )+ tp +
 plot_spacer() + 

    
 plot_annotation(tag_levels = "A", title="Figure 4", theme = theme(title=element_text(size = 20)))+
  plot_layout(design = design1, heights = c(.8,1,1.1), width=c(1.5,1,1))


Fig4

```

## Appendix
```{r appendix4}
Sys.info()
sessionInfo()
```